---
title: "Class08"
author: "Jelani Lyda"
date: "10/25/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## K-means example

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)

```

```{r}
k <- kmeans(x, 2, nstart = 20)
k

```
Q. How many points are in each cluster?
```{r}
#Finds how many points are in each cluster
k$size
```


Q. What ‘component’ of your result object details
 - cluster size?
 - cluster assignment/membership?
 - cluster center?
```{r}
#Cluster assignment/membership
table(k$cluster)
k$centers
```

Plot x colored by the kmeans cluster assignment and
 add cluster centers as blue points
```{r}
plot(x, col = k$cluster)
points(k$centers, col = "blue", pch = 15)
```


##Hierachical clustering in R
```{r}
hc <- hclust( dist(x))
hc
```


Plot my results
```{r}
plot(hc)
abline(h = 6, col = "red")
abline(h = 4, col = "blue")
#Cuts graph at the hight
cutree(hc, h = 6)
grps <- cutree(hc, h = 4)
table(grps)
```

I can also cut the tree to yield a given 'k' groups/clusters 
```{r}
grps2 <- cutree(hc, k=2)
```


```{r}
#plot
plot(x, col = grps2)
```


```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")

plot(x)

# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```

Q. Use the dist(), hclust(), plot() and cutree()
 functions to return 2 and 3 clusters 
```{r}
hc2 <- hclust( dist(x))
plot(hc2)

hc2_k2 <- cutree(hc2, k = 2)
hc2_k3 <- cutree(hc2, k = 3)
```
 
```{r}
plot(x, col = hc2_k2)
plot(x, col = hc2_k3)
```

```{r}
table(col, hc2_k3)
```


```{r}
## You can also download this file from the class website!
mydata <- read.csv("https://tinyurl.com/expression-CSV",
 row.names=1)

head(mydata)
```

```{r}
## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE)
## See what is returned by the prcomp() function
attributes(pca)
```


```{r}
x <- pca$x[,1]
y <- pca$x[,2]
plot(x, y)
plot(pca)
```

```{r}
summary(pca)
```

```{r}
plot(pca$x[,1], pca$x[,2], col = c(rep("red", 5), rep("blue", 5)))
plot(pca$x)
pca$x
```


```{r}
x <- read.csv("UK_foods.csv",
 row.names=1)
dim(x)
head(x)
```
```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
pairs(x, col=rainbow(10), pch=16)
```

```{r}
pca <- prcomp( t(x) )
summary(pca)

plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500), col = c("Orange", "Pink", "Blue", "Green"))
text(pca$x[,1], pca$x[,2], colnames(x), col = c("Orange", "Pink", "Blue", "Green"))
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```










