---
title: "Class14"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup Bioconductor DESeq2

```{r eval=FALSE}
#eval=FALSE: R will ask to run this section of code

install.packages("BiocManager")
BiocManager::install()
# For this class, you'll also need DESeq2:
BiocManager::install("DESeq2")
```

```{r}
library(BiocManager)
```

# Data for todays class
```{r}
counts <- read.csv("airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <- read.csv("airway_metadata.csv", stringsAsFactors = FALSE)
```

```{r}
head(counts)
```

How many genes do we have in the dataset?
```{r}
nrow(counts)
```

How many experiemnts are in this dataset?
```{r}
ncol(counts) - 1
```

Lets make sure the metadata id col matches the colnames of counts
```{r}
#The all function is useful for looking accross a vector of logicals
all(colnames(counts)[-1] == metadata$id)
```

## Analysis: compare the control to the drug treated

First we need to access the columns of out countData that are contol and then find their mean

```{r}
metadata$dex == "control"
```
```{r}
#Get the control ids
control.id <- metadata[metadata$dex == "control", ]$id
control.id
```
```{r}
#Get the control data
head(counts[, control.id])
```

```{r}
#Get the mean of each row
control.mean <- rowSums(counts[,control.id])/length(control.id)
names(control.mean) <- counts$ensgene
```

```{r}
#Now for the treated columns
treated.id <- metadata[metadata$dex == "treated", ]$id
treated.mean <- rowSums(counts[,treated.id])/length(treated.id)
names(treated.mean) <- counts$ensgene
```

Let's store these together
```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

```{r}
#A lot of data are concentrated around 0, so use a log scale to see all of the data
plot(meancounts$control.mean, meancounts$treated.mean, log = "xy")
```

```{r}
#devide by the log of each. If the  value is above 0 then the gene was more expressed in the treatment and if its below 0 then the gene is more expresses in the control
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```
There are a couple of “weird” results. Namely, the NaN ("not a number") and -Inf (negative infinity) results.

The NaN is returned when you divide by zero and try to take the log. The -Inf is returned when you try to take the log of zero. It turns out that there are a lot of genes with zero expression. Let’s filter our data to remove these genes. Again inspect your result (and the intermediate steps) to see if things make sense to you

```{r}
#Test for finding ero entries

x <- c(5,8,0,5)
x == 0
which(x == 0)
```
```{r}
x <- data.frame(happy = c(5,6,0,0), sad = c(0,5,5,0))
x == 0
```
```{r}
#Shows the row and column of TRUE values
which(x == 0, arr.ind = TRUE)
```

Focus on the rows only that have 0 entries
```{r}
unique(which(x == 0, arr.ind = TRUE)[,1])

```

```{r}
a <- data.frame(wanna = c(3,6,54,2,4,0,0), sleep = c(7,0,2,0,1,0,6))
which(a == 0, arr.ind = TRUE)
```

```{r}
#Apply to mean counts
to.rm <- unique(which(meancounts[,1:2] == 0, arr.ind = TRUE)[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.
```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
sum(up.ind)
sum(down.ind)
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```
This is the organism annotation package ("org") for Homo sapiens ("Hs"), organized as an AnnotationDbi database package ("db"), using Entrez Gene IDs ("eg") as primary key. To get a list of all available key types, use:
```{r}
columns(org.Hs.eg.db)
```

We can use the mapIds() function to add individual columns to our results table. We provide the row names of our results table as a key, and specify that keytype=ENSEMBL. The column argument tells the mapIds() function which information we want, and the multiVals argument tells the function what to do if there are multiple possible values for a single input value. Here we ask to just give us back the first one that occurs in the database.
```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts), # Our genenames
                     keytype="ENSEMBL",        # The format of our genenames
                     column="SYMBOL",          # The new format we want to add
                     multiVals="first")
```

```{r}
head(mycounts)
```


Now to finnaly use DESeq2
```{r}
library(DESeq2)
citation("DESeq2")
```

Set up data for DESeq2 analysis
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

Run DESeq
```{r}
dds <- DESeq(dds)
```


```{r}
res <- results(dds)
res
```

```{r}
plot(res$log2FoldChange, res$padj)
```

```{r}
plot(res$log2FoldChange, -log(res$padj))
```
Add colors
```{r}
mycols <- rep("gray", length(res$padj))
mycols[ abs(res$log2FoldChange) > 2 ]  <- "blue"
inds <- (res$padj < 0.01) & (abs(res$log2FoldChange) > 2 )
mycols[ inds ] <- "red"

plot(res$log2FoldChange, -log(res$padj), col = mycols)
abline(v=c(-2,2), col="gray", lty=2)
abline(h=-log(0.1), col="gray", lty=2)
```











